Formalizing the Edmonds-Karp Algorithm

### 2016-03-05 paper/root.tex
Added missing snippets, latex fine-tuning, makefile-setup for snippets
The latex fine tuning is done via a sed-script that processes the generated snippet file. This is a hack, and can be achieved in a cleaner way on isabelle level, by defining output syntax, see Library/LaTeXSugar for examples.

The makefile intentionally does not contain a dependence of the snippets to the Isabelle-theories, to keep it decoupled. To force re-buld of snippets do a
  make clean_snippets
first.
Fixed size problems with listings, restored orig style
2nd iteration over paper
changed presentation of networks and f

@@

@@

\input{lstisabelle}
\lstset{basicstyle=\small\ttfamily\slshape,breaklines=true}
\lstset{captionpos=b}
\lstset{numberbychapter=false}

% Include snippets
\newcommand{\DefineSnippet}[2]{%
   \expandafter\newcommand\csname snippet--#1\endcsname{%
     \begin{quote}
     \begin{isabelle}\footnotesize
     #2
     \end{isabelle}
     \end{quote}}}
\newcommand{\Snippet}[1]{\csname snippet--#1\endcsname}
\input{snippets/snippets}

@@

\maketitle
\begin{abstract}
We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
Our formal proof closely follows a standard textbook proof, and is easy to read even without being
an expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
Further refinement yields a verified implementation, which compares well to an unverified reference implementation in Java.

@@

\section{Introduction}
Computing the maximum flow of a network is an important problem in graph theory.
Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, and various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
which solves the maximum flow problem in polynomial time for the general case of real valued capacities.

In this paper, we present a verified efficient implementation of the Edmonds-Karp algorithm and its polynomial complexity bound.
The formalization is conducted entirely in the Isabelle/HOL proof assistant~\cite{NPW02}. 
Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification into an abstract proof of the Ford-Fulkerson method,
its instantiation to Edmonds-Karp algorithm, and finally its efficient implementation. The abstract parts of our verification closely follow the standard textbook presentation of Cormen et al.~\cite{CLRS09}. Being developed in the Isar~\cite{Wenzel99} proof language, our proofs are accessible even to non-Isabelle experts.

The rest of this paper is structured as follows: In Section~\ref{sec:background} we give a short informal introduction to the Ford-Fulkerson method.
In Section~\ref{sec:abs-formalization}, we report on our formalization of the abstract method. 
Section~\ref{sec:refinement} gives a brief overview of the Isabelle Refinement Framework~\cite{LaTu12,La12}, which supports stepwise refinement based algorithm development in Isabelle/HOL. In Section~\ref{sec:edka}, we report on our instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm and the proof of its complexity.
Section~\ref{sec:executable} reports on the further refinement steps required to yield an efficient implementation. Section~\ref{sec:benchmark} reports on 
benchmarking our implementation against a reference implementation of the Edmonds-Karp algorithm from Sedgewick et al.~\cite{SeWa11}. 
Finally, Section~\ref{sec:concl} gives a conclusion and discusses related and future work.

@@

\section{A short background on the Ford-Fulkerson method}\label{sec:background}
% \begin{comment}
%   GOAL: 
%     1) Remind the reader of FoFu. The educated reader should have the feeling:
%       Yes, now I recall FoFu, and understand (again) what it does.
%       
%     2) Set the field for further, more detailed descriptions in next section.
%   
% 
%   Short description of Network: Finite graph over nodes V, 
%     edge (u,v) annotated with capacity c(u,v).
%     Assuming distinct nodes s and t. 
%     Note on additional assumptions: 
%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
%         Can transform any Network to match these assumptions, preserving the flow \cite{???}.
%         
%   s-t Flow: Annotation of edges with values, such that: 
%     1) Values smaller than capacities.
%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
%   
%   Value of flow: Incoming flow - outgoing flow of s
%   
%   Max-Flow problem. 
%     Rpt. why important?
%   
%   Cuts, intuition of min-cut >= max-flow
%   
%   Theorem: Min-Cut = Max-Flow
%   Proven via augmenting flow in residual graph. 
%     Present 1,2,3 of min-cut max-flow equivalences
%   
%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
%       Termination? In General: Only for integer capacities.
%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
%   \end{comment}

@@

A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real valued capacity $c(u,v)>0$.
Moreover, their are two distinct vertices $s,t\in E$ which are called \emph{source} and \emph{sink}. To simplify reasoning about flow networks, it is common to make some additional assumptions: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no edge $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$. Although these assumptions seem restrictive, any network may be transformed such that it satisfies all the aforementioned properties~\cite{CLRS09}.

An \emph{$s$-$t$ flow} $f$ on a graph is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges. In context of a network, a \emph{flow} is an $s$-$t$ flow, \ie a flow from the source to the sink.

The value of an $s$-$t$ flow $f$ is denoted by $|f|$, and defined to be the sum over all flows leaving $s$ minus the sum over all incoming flows to $s$\footnote{In the context of networks, there are no incoming edges to $s$, such the definition can be simplified.}.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

An important result is the relation between flows and cuts in a network. A \emph{cut} in a network is a partition of the vertices into two sets, such that one set contains 
the source and the other set contains the sink. The capacity of a cut is the sum of the capacities of all edges going from the source's side to the sink's side of the cut.
It is easy to see that the value of any flow cannot exceed the capacity of any cut, as all flow from the source must ultimately reach the sink, and thus go through the edges of the cut. The Ford-Fulkerson theorem tightens this bound and states that the value of the maximum flow is equal to the capacity of the minimum cut.

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose the Ford-Fulkerson method defines the residual graph, which has edges in parallel and reverse to the edges of the network. 
Each edge indicates how much flow can be moved between two nodes by either increasing or decreasing the flow on an edge in the network. Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph, and then 
pushes as much flow as possible along this path to increase the value of the current flow. 
Formally, for an augmenting path $p$, one defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
\[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
An augmenting path yields a residual flow $f_p$, which is the flow that can be passed along this path:
\[ f_p (u, v) = 
  \begin{cases}
  c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
\newcommand{\augment}{{\mathbin\uparrow}}
In order to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
by any \emph{augmenting flow} $f'$, i.e., any flow in the residual graph:
\[ (f \augment f') (u, v) = 
  \begin{cases}
  f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
Note that, for any edge in the network, the augmenting flow parallel to that edge is added to the flow, while the augmenting flow reverse to this edge is subtracted. 
This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.

@@

The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph, and there are several possible implementations with different execution times. The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}. When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$\footnote{Where unambiguous, we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges.}, even for the general case of real valued capacities. A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.

@@

% (DONE: not sure about repeating citation of Cormen: I would definitely repeat the citation!) 
Informal proofs focus on the relevant thoughts by leaving out technical details and obvious steps. 
In contrast, a formal proof has to precisely specify each step as the application of some inference rules. 
Although modern proof assistants provide high-level tactics to summarize some of these steps, formal proofs tend to be significantly more 
verbose than informal proofs. Moreover, formal proofs are conducted in the tactic language of the proof assistant, which is often some dialect of ML. 
Thus, many formal proofs are essentially programs that instruct the proof assistant how to conduct the proof, and are inaccessible without a deep knowledge of the used proof assistant, in many cases requiring to replay the proof in the proof assistant in order to understand the idea behind it.

For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that strongly resemble standard mathematical textbook proofs, and are accessible even for those not familiar with Isabelle/HOL. 
We use Isar to present our proof of the Ford-Fulkerson method such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.

As an example consider the proof that augmenting a flow $f$ with a residual flow $f'$ yields a valid flow $f\augment f'$ again. In particular, one has to show 
that the augmented flow satisfies the capacity constraint. Cormen et al.~give the following proof, which we copied literally, only replacing the references to 
``Equation~26.4'' by ``definition of $\augment$'':

For the capacity constraint, first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of $\augment$)} \\
	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
	& = f'(u, v) \\
	& \geq 0.
	\end{align*}
In addition,
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(by definition of $\augment$)} \\
	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
	& \leq f(u, v) + c_f(u, v) &&  \text{(capacity constraint)} \\
	& = f(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
	& = c (u, v).
	\end{align*}

@@

TODO: Introduce obtain-syntax, and unfold find-augmenting-path spec! Or even while-exists loop as combinator?
The code looks quite similar to pseudocode that one would expect in a textbook, but actually is a rigorous formal specification of the algorithm, using nondeterminism to leave open the implementation details (\cf Section~\ref{sec:refinement}). Note that we had to use the available combinators of the refinement framework, which made the 
code slightly more verbose than we would have liked.

@@

Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Refinement Framework, it is straightforward to prove (partial) correctness of
the Ford-Fulkerson method, which is stated in Isabelle/HOL by the following theorem:
\Snippet{ford_fulkerson_partial_correct}

\section{Refinement in Isabelle/HOL}\label{sec:refinement}
After having stated and proved correct an algorithm on the abstract level, the next step is to provide an efficient implementation. 
In our case, we first specialize the Ford-Fulkerson method to use shortest augmenting paths, then implement the search for shortest augmenting paths by BFS, and finally use efficient data structures to represent the abstract objects modified by the algorithm.

@@

In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12}. 
It features a refinement calculus for programs phrased in a nondeterminism monad. 
The monad's type is a set of possible results plus an additional value that indicates a failure:
\begin{lstlisting}
  datatype \<alpha> nres = res \<alpha> set | fail
\end{lstlisting}
The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
operation \isai{bind m f} nondeterministically picks a result from $m$ and executes $f$ on it. 
If either \isai{m = fail}, or \isai{f} may fail for a result in $m$, the bind operation fails.

We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
Intuitively, $m\le m'$ means that $m$ is a refinement of $m'$, i.e., all possible results of $m$ are 
also possible results of $m'$. 
Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define constants for loop constructs (while, foreach), 
assertions and specifications, and use a Haskell-like do-notation:
\begin{lstlisting}
  assert \<Phi> = if \<Phi> then return () else fail
  spec x. P x = RES {x. P x}
  do {x <- m; f x} = bind m f
  do {m; m'} = bind m (%_. m')
\end{lstlisting}

Correctness of a program $m$ with precondition $P$ and postcondition $Q$ is expressed as \isai{P ==> m <= spec r. Q r} (or, eta-contracted, just \isai{spec Q}), which
means that, if $P$ holds, $m$ does not fail and all possible results of $m$ satisfy $Q$. Note that we provide different recursion constructs
for partial and total correctness: A nonterminating total correct recursion yields \isai{fail}, while a nonterminating partial correct 
recursion yields \isai|res {}|, which is the least element of the refinement ordering.

@@

In a typical program development, one first comes up with an initial version $m_0$ of
the algorithm and its specification $P,Q$, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions $m_i$ of the algorithm,
proving \isai{$m_i$ <= \<Down>$R_i$ $m_{i-1}$}. Using transitivity and composability of data refinement, one 
gets \isai{P ==> $m_i$ <= \<Down>$R_i\ldots R_1$ spec Q}, showing the correctness of the refined algorithm.
Various tools, including a verification condition generator, assist the use in conducting the refinement proofs by breaking 
them down to statements that do not involve monad constructs any more, and in many cases precisely reflect the core idea of the proof.

@@

\section{Edmonds-Karp Algorithm}\label{sec:edka}
  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm requires us to show that finding a shortest augmenting path is a correct refinement of finding any augmenting path, which is straightforward using basic graph theory.

  Considerable more effort is required on showing that the resulting algorithm terminates within $O(VE)$ iterations. 
  The idea of the proof is as follows: Edges that are reverse to edges on a shortest path cannot lie 
  on a shortest path itself.
  On every augmentation, at least one edge of the residual graph that lies on a shortest augmenting path is flipped. Thus, either the length of the shortest path increases,
  or the number of edges that lie on a shortest path decreases. As the length of a shortest path is at most $V$, there are no more than $O(VE)$ iterations.

@@

  Formalizing the above intuitive argument was more tricky than it seemed on first glance:
  While it easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
  to a graph transformation which may add reverse edges and removes at least one original edge requires some generalization of the statement.
  Note that a straightforward induction on the length of the augmenting path or on the number of flipped edges fails, as, after flipping the first 
  edge, the path is no longer augmenting.

@@

    %start data for plotting
      \addplot table {
        1000 204
        2000 1430
        3000 4928
        4000 9299
        5000 23625
      }; \addlegendentry{SML, Sparse};

>>>>

% \isabellestyle{it}

\def\isachardoublequote{}%
\def\isachardoublequoteopen{}%
\def\isachardoublequoteclose{}%

\newcommand{\isainnerkeyword}[1]{{\tt #1}}
\newcommand{\isasymexistsA}{\isamath{\exists_{\textsc A}\,}}

@@

\input{lstisabelle}
\lstset{basicstyle=\footnotesize\ttfamily\slshape}
\lstset{captionpos=b}
\lstset{numberbychapter=false}

\newcommand{\isai}{\lstinline[language=isabelle,basicstyle=\normalsize\ttfamily\slshape]}

@@

% Include snippets
\newcommand{\DefineSnippet}[2]{%
   \expandafter\newcommand\csname snippet--#1\endcsname{%
     \begin{quote}
     \begin{isabelle}\footnotesize
     #2
     \end{isabelle}
     \end{quote}}}
\newcommand{\Snippet}[1]{\ifcsname snippet--#1\endcsname\csname snippet--#1\endcsname\else\PackageError{}{No snippet '#1' defined.}{}\fi}
\input{snippets/snippets.out}

@@

\maketitle
\begin{abstract}
We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
Our formal proof closely follows a standard textbook proof, and is accessible even without being
an expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
Further refinement yields a verified implementation, whose execution time compares well to an unverified reference implementation in Java.

@@

\section{Introduction}
Computing the maximum flow of a network is an important problem in graph theory.
Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, and various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
which was one of the first algorithms to solve the maximum flow problem in polynomial time for the general case of real valued capacities.

In this paper, we present a formal verification of the Edmonds-Karp algorithm and its polynomial complexity bound.
The formalization is conducted entirely in the Isabelle/HOL proof assistant~\cite{NPW02}. 
Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification into an abstract proof of the Ford-Fulkerson method,
its instantiation to Edmonds-Karp algorithm, and finally its efficient implementation. The abstract parts of our verification closely follow the standard textbook presentation of Cormen et al.~\cite{CLRS09}. Being developed in the Isar~\cite{Wenzel99} proof language, our proofs are accessible even to non-Isabelle experts.

While there exists another formalization of the Ford-Fulkerson method~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation.

The rest of this paper is structured as follows: In Section~\ref{sec:background} we give a short informal introduction to the Ford-Fulkerson method.
In Section~\ref{sec:abs-formalization}, we report on our formalization of the abstract method. 
Section~\ref{sec:refinement} gives a brief overview of the Isabelle Refinement Framework~\cite{LaTu12,La12}, which supports stepwise refinement based algorithm development in Isabelle/HOL. In Section~\ref{sec:edka}, we report on our instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm and the proof of its complexity.
Section~\ref{sec:executable} reports on the further refinement steps required to yield an efficient implementation. Section~\ref{sec:benchmark} reports on 
benchmarking our implementation against a reference implementation of the Edmonds-Karp algorithm from Sedgewick et al.~\cite{SeWa11}. 
Finally, Section~\ref{sec:concl} gives a conclusion and discusses related and future work. The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/edmonds_karp/}.

@@

\section{A short background on the Ford-Fulkerson method}\label{sec:background}
% \begin{comment}
%   GOAL: 
%     1) Remind the reader of FoFu. The educated reader should have the feeling:
%       Yes, now I recall FoFu, and understand (again) what it does.
%       
%     2) Set the field for further, more detailed descriptions in next section.
%   
% 
%   Short description of Network: Finite graph over nodes V, 
%     edge (u,v) annotated with capacity c(u,v).
%     Assuming distinct nodes s and t. 
%     Note on additional assumptions: 
%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
%         Can transform any Network to match these assumptions, preserving the flow cite{???}.
%         
%   s-t Flow: Annotation of edges with values, such that: 
%     1) Values smaller than capacities.
%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
%   
%   Value of flow: Incoming flow - outgoing flow of s
%   
%   Max-Flow problem. 
%     Rpt. why important?
%   
%   Cuts, intuition of min-cut >= max-flow
%   
%   Theorem: Min-Cut = Max-Flow
%   Proven via augmenting flow in residual graph. 
%     Present 1,2,3 of min-cut max-flow equivalences
%   
%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
%       Termination? In General: Only for integer capacities.
%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
%   \end{comment}

@@

% TODO: Possible alternative presentation
%   First define a relaxed network, without the constraints. Define the flow there. Then add network constraints.
%

A (flow) network is a directed graph over a finite set of vertices $V$ and edges $E$, where each edge $(u,v)\in E$ is labeled by a positive real valued capacity $c(u,v)>0$.
Moreover, there are two distinct vertices $s,t\in V$ which are called \emph{source} and \emph{sink}.

An \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges. 
The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$. Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.

An important result is the relation between flows and cuts in a network. A \emph{cut} is a partitioning of the vertices into two sets, such that one set contains 
the source and the other set contains the sink. The capacity of a cut is the sum of the capacities of all edges going from the source's side to the sink's side of the cut.
It is easy to see that the value of any flow cannot exceed the capacity of any cut, as all flow from the source must ultimately reach the sink, and thus go through the edges of the cut. The Ford-Fulkerson theorem tightens this bound and states that the value of the maximum flow is equal to the capacity of the minimum cut.

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose the Ford-Fulkerson method defines the residual graph, which has edges both in parallel and reverse to the edges of the network. 
Each edge is labeled by the amount of flow that can be effectively moved along this edge, by either increasing or decreasing the flow on an edge in the network.
Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph, and then 
pushes as much flow as possible along this path to increase the value of the current flow. 
Formally, for an augmenting path $p$, one defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
\[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
An augmenting path yields a residual flow $f_p$, which is the flow that can be passed along this path:
\[ f_p (u, v) = 
  \begin{cases}
  c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
\newcommand{\augment}{{\mathbin\uparrow}}
In order to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
by any \emph{augmenting flow} $f'$, \ie any flow in the residual graph:
\[ (f \augment f') (u, v) = 
  \begin{cases}
  f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
Note that, for any edge in the network, the augmenting flow parallel to that edge is added to the flow, while the augmenting flow reverse to this edge is subtracted. 
This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.

@@

The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph, and there are several possible implementations with different execution times. The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}. When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real valued capacities. Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges.
A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.

@@

% (DONE: not sure about repeating citation of Cormen: I would definitely repeat the citation!) 
Informal proofs focus on the relevant thoughts by leaving out technical details and obvious steps. 
In contrast, a formal proof has to precisely specify each step as the application of some inference rules. 
Although modern proof assistants provide high-level tactics to summarize some of these steps, formal proofs tend to be significantly more 
verbose than informal proofs. Moreover, formal proofs are conducted in the tactic language of the proof assistant, which is often some dialect of ML. 
Thus, many formal proofs are essentially programs that instruct the proof assistant how to conduct the proof. They tend to be inaccessible without a deep knowledge of the used proof assistant, in many cases requiring to replay the proof in the proof assistant in order to understand the idea behind it.

For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that resemble standard mathematical textbook proofs, and are accessible, to a certain extent, even for those not familiar with Isabelle/HOL. 
We use Isar to present our proof of the Ford-Fulkerson method such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.

As an example consider the proof that augmenting a flow $f$ with a residual flow $f'$ yields a valid flow $f\augment f'$ again. In particular, one has to show 
that the augmented flow satisfies the capacity constraint. Cormen et al.~give the following proof, which we display literally here, only replacing the references to 
``Equation~26.4'' by ``definition of $\augment$'':

For the capacity constraint, first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
	& = f'(u, v) \\
	& \geq 0.
	\end{align*}
In addition,
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
	& \leq f(u, v) + c_f(u, v) &&  \text{(capacity constraint)} \\
	& = f(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
	& = c (u, v).
	\end{align*}

@@

% DONE: Introduce obtain-syntax, and unfold find-augmenting-path spec! Or even while-exists loop as combinator?
The code looks quite similar to pseudocode that one would expect in a textbook, but actually is a rigorous formal specification of the algorithm, using nondeterminism to leave open the implementation details (\cf Section~\ref{sec:refinement}). Note that we had to use the available combinators of the Isabelle Refinement Framework, which made the code slightly more verbose than we would have liked. We leave it to future work to define a set of combinators and appropriate syntax that allow for more concise presentation of pseudocode.

@@

Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Refinement Framework, it is straightforward to prove (partial) correctness of
the Ford-Fulkerson method, which is stated in Isabelle/HOL by the following theorem:
\Snippet{ford_fulkerson_correct}

\section{Refinement in Isabelle/HOL}\label{sec:refinement}
After having stated and proved correct an algorithm on the abstract level, the next step is to provide an (efficient) implementation. 
In our case, we first specialize the Ford-Fulkerson method to use shortest augmenting paths, then implement the search for shortest augmenting paths by BFS, and finally use efficient data structures to represent the abstract objects modified by the algorithm.

@@

In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12,La12}. 
It features a refinement calculus for programs phrased in a nondeterminism monad. 
The monad's type is a set of possible results plus an additional value that indicates a failure:
\begin{lstlisting}
  datatype \<alpha> nres = res \<alpha> set | fail
\end{lstlisting}
The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
operation \isai{bind m f} nondeterministically picks a result from $m$ and executes $f$ on it. 
If either \isai{m = fail}, or \isai{f} may fail for a result in $m$, the bind operation fails.

We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
Intuitively, $m\le m'$ means that $m$ is a refinement of $m'$, \ie all possible results of $m$ are 
also possible results of $m'$. 
Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define standard loop constructs (while, foreach), 
a syntax for postcondition specifications, and use a Haskell-like do-notation:
\begin{lstlisting}
  spec P == spec x. P x == res {x. P x}
  do {x <- m; f x} == bind m f
  do {m; m'} == bind m (%_. m')
\end{lstlisting}

Correctness of a program $m$ with precondition $P$ and postcondition $Q$ is expressed as \isai{P ==> m <= spec r. Q r} (or, eta-contracted, just \isai{spec Q}), which
means that, if $P$ holds, $m$ does not fail and all possible results of $m$ satisfy $Q$. Note that we provide different recursion constructs
for partial and total correctness: A nonterminating total correct recursion yields \isai{fail}, which satisfies no specification, even if joined 
with results from other possible runs. On the other hand, a nonterminating partial correct recursion yields \isai|res {}|, which refines any specification 
and disappears when joined with other results.

@@

In a typical program development, one first comes up with an initial version $m_0$ of
the algorithm and its specification $P,Q$, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions $m_i$ of the algorithm,
proving \isai{$m_i$ <= \<Down>$R_i$ $m_{i-1}$}. Using transitivity and composability of data refinement, one 
gets \isai{P ==> $m_i$ <= \<Down>$R_i\ldots R_1$ spec Q}, showing the correctness of the refined algorithm.
If no data refinement is performed, $R_i$ is set to the identity relation, in which case \isai{\<Down>$R_i$} becomes the identity function.

Various tools, including a verification condition generator, assist the user in conducting the refinement proofs by breaking 
them down to statements that do not involve monad constructs any more, and in many cases precisely reflect the core idea of the proof.

@@

\section{The Edmonds-Karp Algorithm}\label{sec:edka}
  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm requires us to show that finding a shortest augmenting path is a refinement of finding any augmenting path, which is straightforward using basic graph theory.

  Considerably more effort is required to show that the resulting algorithm terminates within $O(VE)$ iterations. 
  The idea of the proof is as follows: Edges that are reverse to edges on a shortest path cannot lie 
  on a shortest path itself.
  On every augmentation, at least one edge of the residual graph that lies on a shortest augmenting path is flipped. Thus, either the length of the shortest path increases,
  or the number of edges that lie on a shortest path decreases. As the length of a shortest path is at most $V$, there are no more than $O(VE)$ iterations.

@@

  Formalizing the above intuitive argument was more tricky than it seemed on first glance:
  While it is easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
  to a graph transformation which may add reverse edges and removes at least one original edge requires some generalization of the statement.
  Note that a straightforward induction on the length of the augmenting path or on the number of flipped edges fails, as, after flipping the first 
  edge, the path is no longer augmenting.

@@

    %start data for plotting
      \addplot table {
        1000 204
        2000 1430
        3000 4928
        4000 9299
        5000 23625
      }; \addlegendentry{SML, sparse};
